---
title: "MLProj"
author: "Pradeep Gurav"
date: "23/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
The goal of this project is to predict the manner in which subjects did the exercise. The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. I have used RandomForest method to build the model. This report describes:  
* how the model is built  
* use of cross validation  
* an estimate of expected out of sample error  
* Predictied values for the Testdata provided



## Getting and cleaning the Data
```{r include=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(e1071)
library(randomForest)
```

```{r }
set.seed(123)

train.url <-
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url <- 
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

path <- paste(getwd(),"/", sep="")
train.file <- file.path(path, "machine-train-data.csv")
test.file <- file.path(path, "machine-test-data.csv")
if (!file.exists(train.file)) {
        download.file(train.url, destfile=train.file)
}
if (!file.exists(test.file)) {
        download.file(test.url, destfile=test.file)
}

train.data.raw <- read.csv(train.file, na.strings=c("NA","#DIV/0!",""))
test.data.raw <- read.csv(test.file, na.strings=c("NA","#DIV/0!",""))

## Remove irrelevant colums

# Drop the first 7 columns as they're not relevant for predicting.
train_data <- train.data.raw[,8:length(colnames(train.data.raw))]
test_data <- test.data.raw[,8:length(colnames(test.data.raw))]

# Drop colums with NAs
train_data <- train_data[, colSums(is.na(train_data)) == 0] 
test_data <- test_data[, colSums(is.na(test_data)) == 0] 

# Check for near zero variance predictors and drop them if necessary
nzv <- nearZeroVar(train_data,saveMetrics=TRUE)
zero.var.ind <- sum(nzv$nzv)

if ((zero.var.ind>0)) {
        train_data <- train_data[,nzv$nzv==FALSE]
}

# Check for near zero variance predictors and drop them if necessary



```

## Split the data for cross validation
The training data is divided into two sets. This first is a training set with 70% of the data which is used to train the model. The second is a cross validation set used to assess model performance.

```{r }
in.training <- createDataPartition(train_data$classe, p=0.70, list=F)
train.data.final <- train_data[in.training, ]
crossvalidata <- train_data[-in.training, ]
```

## Model Development

We will use random forest as the model as implemented in the randomForest package.

# Why we will use RandomForest method to build a model

Because  
* it automatically selects important variables and  
* is robust to correlated covariates & outliers in general  
* 5-fold cross validation is used in the algorithm.  
* averages multiple deep decision trees  
 

```{r }
control.parms <- trainControl(method="cv", 5)
rf.model <- train(classe ~ ., data=train.data.final, method="rf",
                 trControl=control.parms, ntree=100)

# Training set accuracy
#ptraining <- predict(rf.model, train.data.final)
#print(confusionMatrix(ptraining, train.data.final$classe))
```

Obviously the model performs excellent against the training set, but we need to cross validate the performance against the held out set and see if we have avoided overfitting.

## Cross Validation using the Validation dataset (Out of Sample)
Let us now see how the model performs on the cross validation set that we held out from training.
```{r }
rf.predict <- predict(rf.model, crossvalidata)
print(confusionMatrix(crossvalidata$classe, rf.predict))
```

The cross validation accuracy is 99.32% and the out-of-sample error is therefore 0.68% so the model performs rather good.

# Out of sample error with the model is .68%

## Test set prediction
The prediction of the algorithm for the test set is:

```{r }
results <- predict(rf.model, 
                   test_data[, -length(names(test_data))])
results
```

We then save the output to files according to instructions and post it to the submission page.


## Reference
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13). Stuttgart, Germany: ACM SIGCHI, 2013

## Annexure Graph
```{r }
ImpObj <- varImp(rf.model)
plot(ImpObj, main = "Top 25 influencing Variables", top = 25)
```