---
title: "MLProj"
author: "Pradeep Gurav"
date: "21/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
The goal of this project is to predict the manner in which subjects did the exercise. The model will use the other variables to predict with. This report describes:  
* how the model is built  
* use of cross validation  
* an estimate of expected out of sample error  


## Getting and cleaning the Data
```{r include=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(e1071)
library(randomForest)
```

```{r }
set.seed(123)

train.url <-
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url <- 
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

path <- paste(getwd(),"/", sep="")
train.file <- file.path(path, "machine-train-data.csv")
test.file <- file.path(path, "machine-test-data.csv")
if (!file.exists(train.file)) {
        download.file(train.url, destfile=train.file)
}
if (!file.exists(test.file)) {
        download.file(test.url, destfile=test.file)
}

train.data.raw <- read.csv(train.file, na.strings=c("NA","#DIV/0!",""))
test.data.raw <- read.csv(test.file, na.strings=c("NA","#DIV/0!",""))

## Remove irrelevant colums

# Drop the first 7 columns as they're not relevant for predicting.
train_data <- train.data.raw[,8:length(colnames(train.data.raw))]
test_data <- test.data.raw[,8:length(colnames(test.data.raw))]

# Drop colums with NAs
train_data <- train_data[, colSums(is.na(train_data)) == 0] 
test_data <- test_data[, colSums(is.na(test_data)) == 0] 

# Check for near zero variance predictors and drop them if necessary
nzv <- nearZeroVar(train_data,saveMetrics=TRUE)
zero.var.ind <- sum(nzv$nzv)

if ((zero.var.ind>0)) {
        train_data <- train_data[,nzv$nzv==FALSE]
}
```

## Split the data for cross validation
The training data is divided into two sets. This first is a training set with 70% of the data which is used to train the model. The second is a cross validation set used to assess model performance.

```{r }
in.training <- createDataPartition(train_data$classe, p=0.70, list=F)
train.data.final <- train_data[in.training, ]
crossvalidata <- train_data[-in.training, ]
```

## Model Development

We will use random forest as the model as implemented in the randomForest package.

# Why we will use RandomForest method to build a model

Because it automatically selects important variables and is robust to correlated covariates & outliers in general. 5-fold cross validation is used when applying the algorithm. A Random Forest algorithm is a way of averaging multiple deep decision trees, trained on different parts of the same data-set, with the goal of reducing the variance. This typically produces better performance at the expense of bias and interpret-ability.  
The Cross-validation technique assesses how the results of a statistical analysis will generalize to an independent data set. In 5-fold cross-validation, the original sample is randomly partitioned into 5 equal sized sub-samples. a single sample is retained for validation and the other sub-samples are used as training data. The process is repeated 5 times and the results from the folds are averaged.


```{r }
control.parms <- trainControl(method="cv", 5)
rf.model <- train(classe ~ ., data=train.data.final, method="rf",
                 trControl=control.parms, ntree=251)

# Training set accuracy
ptraining <- predict(rf.model, train.data.final)
print(confusionMatrix(ptraining, train.data.final$classe))
```

Obviously the model performs excellent against the training set, but we need to cross validate the performance against the held out set and see if we have avoided overfitting.

## Cross Validation using the Validation dataset (Out of Sample)
Let us now see how the model performs on the cross validation set that we held out from training.
```{r }
rf.predict <- predict(rf.model, crossvalidata)
print(confusionMatrix(crossvalidata$classe, rf.predict))
```

The cross validation accuracy is 99.34% and the out-of-sample error is therefore 0.67% so the model performs rather good.

# Out of sample error with the model is .67%

## Test set prediction
The prediction of the algorithm for the test set is:

```{r }
results <- predict(rf.model, 
                   test_data[, -length(names(test_data))])
results
```

We then save the output to files according to instructions and post it to the submission page.

```{r }
answers <- as.vector(results)

pml_write_files = function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
            col.names = FALSE)
    }
}

pml_write_files(answers)
```


## Reference
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13). Stuttgart, Germany: ACM SIGCHI, 2013

## Annexure Graph
```{r }
ImpObj <- varImp(rf.model)
plot(ImpObj, main = "Top 25 influencing Variables", top = 25)
```